Continuous gravitational waves in the lab: recovering audio signals with a table-top~optical microphone

### 2021-11-01 paper/ifo-appendix.tex
removing our comments to share source files
changing eqn to eq for AJP style guide

The likelihood of measuring the signal $s(t)$ in the presence of noise follows from Eqns.~\ref{eqn:GNdata} and~\ref{eqn:ipnn} by replacing $n(t)$ in Eqn.~\ref{eqn:ipnn} with $x(t) - s(t) = n(t)$ from Eqn.~\ref{eqn:GNdata}.~\cite{JKS:1998,Jaynes:2003}
The result is
\begin{eqnarray}
\mathcal{L} &=& \exp\left( -\frac{1}{2}\ip{x-s}{x-s} \right) \label{eqn:likeOne}\,, \\ 
            &=& \exp\left( -\frac{1}{2}\ip{x}{x} - \frac{A^2}{2} \ip{\cos[\omega t + \phi]}{\cos[\omega t + \phi]} \right. \nonumber\\
                 &&\hspace{3em} \left. +  A \ip{x}{\cos[\omega t + \phi]} \vphantom{\frac12}\right) \,,\\ 
                 &=& \exp \left( -\frac{1}{2}\ip{x}{x} - \frac{A^2}{4} + A \ip{x}{\cos[\omega t + \phi]} \right)\,.\label{eqn:logLike}
\end{eqnarray}
We then maximise Eqn~.\ref{eqn:logLike} with respect to $A$, obtaining 
\begin{eqnarray}
\label{eqn:maxLogLike}
\mathcal{L}_\mathrm{max} = \exp \left( - \frac{1}{2} \ip{x}{x} + \ip{x}{\cos[\omega t + \phi]}^2 \right)\,,
\end{eqnarray}
for $A = 2\ip{x}{\cos[\omega t + \phi]}$.
From the second term of Eqn.~\ref{eqn:maxLogLike}, we see that the maximum likelihood of a sinusoidal signal in Gaussian noise is the modulus of the cosine Fourier transform, plus the term $\ip{x}{x}$, which is independent of $\omega$ and $\phi$ and can therefore be ignored when searching over $\omega$.

Two important points must be made about the above procedure. 
(i) Fundamentally the goal is to maximize $\mathcal{L}$ in Eqn.~\ref{eqn:likeOne} by varying $s(t)$ through $A$.
For the special case of the Gaussian likelihood (Eqn.~\ref{eqn:ipnn}), which peaks at $\ip{n}{n}= 0$, this is equivalent to minimizing the difference between $x(t)$ and $s(t)$ as evident in Eqn.~\ref{eqn:ipuv}.
In general, however, minimizing the difference between $x(t)$ and $s(t)$ is not equivalent always to the fundamental goal of maximising $\mathcal{L}$, for example if $\mathcal{L}$ peaks at $\ip{n}{n} \neq 0$, or if $\mathcal{L}$ has multiple maxima.
(ii) The maximum likelihood $\mathcal{L}_{\rm max}$ in Eqn.~\ref{eqn:maxLogLike} (or equivalently its logarithm) defines the detection statistic. 
When its value exceeds a threshold (chosen freely by the analyst) at some value of $\omega$, a signal is deemed to have been detected at that value of $\omega$.
Therefore the specific functional form of Eqn.~\ref{eqn:maxLogLike} matters, which is a second reason why one must start from Eqn.~\ref{eqn:ipnn} rather than Eqn.~\ref{eqn:likeOne}, in addition to reason (i).

\section{Viterbi algorithm}
\label{app:viterbi}
This appendix contains some details regarding the implementation of the Viterbi algorithm described in Section~\ref{sec:viterbi}. 
The Viterbi algorithm~\cite{Viterbi:1967} is a classic method in signal processing, whose theoretical underpinnings and implementation are accessible to undergraduate students. 
See the Supplementary Material for further resources and other pseudocode examples available online.

@@

\begin{algorithmic}
\Function{Viterbi}{$X$}
    % \State $a \gets a+1$
    % \State \Return $a$

>>>>

The likelihood of measuring the signal $s(t)$ in the presence of noise follows from Eqs.~\ref{eqn:GNdata} and~\ref{eqn:ipnn} by replacing $n(t)$ in Eq.~\ref{eqn:ipnn} with $x(t) - s(t) = n(t)$ from Eq.~\ref{eqn:GNdata}.~\cite{JKS:1998,Jaynes:2003}
The result is
\begin{eqnarray}
\mathcal{L} &=& \exp\left( -\frac{1}{2}\ip{x-s}{x-s} \right) \label{eqn:likeOne}\,, \\ 
            &=& \exp\left( -\frac{1}{2}\ip{x}{x} - \frac{A^2}{2} \ip{\cos[\omega t + \phi]}{\cos[\omega t + \phi]} \right. \nonumber\\
                 &&\hspace{3em} \left. +  A \ip{x}{\cos[\omega t + \phi]} \vphantom{\frac12}\right) \,,\\ 
                 &=& \exp \left( -\frac{1}{2}\ip{x}{x} - \frac{A^2}{4} + A \ip{x}{\cos[\omega t + \phi]} \right)\,.\label{eqn:logLike}
\end{eqnarray}
We then maximise Eq.~\ref{eqn:logLike} with respect to $A$, obtaining 
\begin{eqnarray}
\label{eqn:maxLogLike}
\mathcal{L}_\mathrm{max} = \exp \left( - \frac{1}{2} \ip{x}{x} + \ip{x}{\cos[\omega t + \phi]}^2 \right)\,,
\end{eqnarray}
for $A = 2\ip{x}{\cos[\omega t + \phi]}$.
From the second term of Eq.~\ref{eqn:maxLogLike}, we see that the maximum likelihood of a sinusoidal signal in Gaussian noise is the modulus of the cosine Fourier transform, plus the term $\ip{x}{x}$, which is independent of $\omega$ and $\phi$ and can therefore be ignored when searching over $\omega$.

Two important points must be made about the above procedure. 
(i) Fundamentally the goal is to maximize $\mathcal{L}$ in Eq.~\ref{eqn:likeOne} by varying $s(t)$ through $A$.
For the special case of the Gaussian likelihood (Eq.~\ref{eqn:ipnn}), which peaks at $\ip{n}{n}= 0$, this is equivalent to minimizing the difference between $x(t)$ and $s(t)$ as evident in Eq.~\ref{eqn:ipuv}.
In general, however, minimizing the difference between $x(t)$ and $s(t)$ is not equivalent always to the fundamental goal of maximising $\mathcal{L}$, for example if $\mathcal{L}$ peaks at $\ip{n}{n} \neq 0$, or if $\mathcal{L}$ has multiple maxima.
(ii) The maximum likelihood $\mathcal{L}_{\rm max}$ in Eq.~\ref{eqn:maxLogLike} (or equivalently its logarithm) defines the detection statistic. 
When its value exceeds a threshold (chosen freely by the analyst) at some value of $\omega$, a signal is deemed to have been detected at that value of $\omega$.
Therefore the specific functional form of Eq.~\ref{eqn:maxLogLike} matters, which is a second reason why one must start from Eq.~\ref{eqn:ipnn} rather than Eq.~\ref{eqn:likeOne}, in addition to reason (i).

\section{Viterbi algorithm}
\label{app:viterbi}

This appendix contains some details regarding the implementation of the Viterbi algorithm described in Section~\ref{sec:viterbi}. 
The Viterbi algorithm~\cite{Viterbi:1967} is a classic method in signal processing, whose theoretical underpinnings and implementation are accessible to undergraduate students. 
See the Supplementary Material for further resources and other pseudocode examples available online.

@@

\begin{algorithmic}
\Function{Viterbi}{$X$}

### 2021-11-01 paper/ifo-supplementary-material.tex
removing our comments to share source files
changing eqn to eq for AJP style guide

\newcommand{\jam}{\textcolor{magenta}}
\newcommand{\han}{\textcolor{orange}}

@@

A general band-pass filter combines a high-pass filter and a low-pass filter to smoothly attenuate frequencies outside of some band (alt.\ pass-band). A Butterworth band-pass filter is a particular band-pass filter such that the frequency response (the attenuation at each frequency) is ``maximally flat'' within the band. The Butterworth low-pass component is given by 
\begin{equation}
\label{eq:butterworth}
H(f) = \left[1+\varepsilon^2 \left( \frac{f}{f_c} \right)^{2n}\right]^{-1/2},
\end{equation}
and is combined with a similar high-pass filter to form the band-pass filter.
In Eqn.~\ref{eq:butterworth}, $f_c$ is the cut-off frequency of the low-pass Butterworth filter, $\varepsilon$ is the gain, and $n$ is the order of the filter which determines how quickly the response rolls off past the cut-off frequency. Fig.~\ref{fig:butterworth} shows the frequency response of the filter used here: a fifth order ($n = 5$) Butterworth filter with a pass-band of $(150\,{\rm Hz}$, $3\,{\rm kHz})$. This high frequency ($3\,{\rm kHz}$) cut-off is chosen since the frequencies important for speech generally lie below $2\,{\rm kHz}$~\cite{speech_intelligibility}.

@@

If we let the Wiener filter be non-causal (i.e. we estimate the current signal based on both past \emph{and future} observations), then we can represent Eqn.~\ref{eqn:wiener-hopf} in the frequency domain as
\begin{equation}
\label{eqn:wiener}
    H(f)=\frac{P_{ss}(f)}{P_{xx}(f)}=\frac{P_{ss}(f)}{P_{ss}(f)+P_{ww}(f)}\,,
\end{equation}
where $P_{xx}(f), P_{ss}(f), P_{ww}(f)$ are the spectra of the observed noisy data, the injected signal, and the background noise, respectively. Intuitively, from Eqn.~\ref{eqn:wiener}, we can see that the non-causal Wiener filter amplifies the input signal where the signal-to-noise ratio (SNR) is high and attenuates the signal where the SNR is low. The causal Wiener filter (which only makes estimates based on past observations) is similar. A more detailed analysis of both kinds of Wiener filters can be found in Ref.~\citep{10.5555/151045}.

In this work, we construct a higher-order causal Wiener filter based on Eqn.~\ref{eqn:wiener-hopf}. A higher-order Wiener filter provides greater smoothing of the input signal but also increases the computational memory required. For this work, we choose a Wiener filter of order $n=100$ as it provides a reasonable balance between smoothing and efficiency. The timeseries and frequency spectrum after applying the Wiener filter are shown in the sixth row in Fig.~\ref{fig:op_mic_results}.
We see a significant improvement in the timeseries of the recovered signal, however, a strong noise hum persists, audibly.

>>>>

@@

A general band-pass filter combines a high-pass filter and a low-pass filter to smoothly attenuate frequencies outside of some band (alt.\ pass-band). A Butterworth band-pass filter is a particular band-pass filter such that the frequency response (the attenuation at each frequency) is ``maximally flat'' within the band. The Butterworth low-pass component is given by 
\begin{equation}
\label{eq:butterworth}
H(f) = \left[1+\varepsilon^2 \left( \frac{f}{f_c} \right)^{2n}\right]^{-1/2},
\end{equation}
and is combined with a similar high-pass filter to form the band-pass filter.
In Eq.~\ref{eq:butterworth}, $f_c$ is the cut-off frequency of the low-pass Butterworth filter, $\varepsilon$ is the gain, and $n$ is the order of the filter which determines how quickly the response rolls off past the cut-off frequency. Fig.~\ref{fig:butterworth} shows the frequency response of the filter used here: a fifth order ($n = 5$) Butterworth filter with a pass-band of $(150\,{\rm Hz}$, $3\,{\rm kHz})$. This high frequency ($3\,{\rm kHz}$) cut-off is chosen since the frequencies important for speech generally lie below $2\,{\rm kHz}$~\cite{speech_intelligibility}.

@@

If we let the Wiener filter be non-causal (i.e. we estimate the current signal based on both past \emph{and future} observations), then we can represent Eq.~\ref{eqn:wiener-hopf} in the frequency domain as
\begin{equation}
\label{eqn:wiener}
    H(f)=\frac{P_{ss}(f)}{P_{xx}(f)}=\frac{P_{ss}(f)}{P_{ss}(f)+P_{ww}(f)}\,,
\end{equation}
where $P_{xx}(f), P_{ss}(f), P_{ww}(f)$ are the spectra of the observed noisy data, the injected signal, and the background noise, respectively. Intuitively, from Eq.~\ref{eqn:wiener}, we can see that the non-causal Wiener filter amplifies the input signal where the signal-to-noise ratio (SNR) is high and attenuates the signal where the SNR is low. The causal Wiener filter (which only makes estimates based on past observations) is similar. A more detailed analysis of both kinds of Wiener filters can be found in Ref.~\citep{10.5555/151045}.

In this work, we construct a higher-order causal Wiener filter based on Eq.~\ref{eqn:wiener-hopf}. A higher-order Wiener filter provides greater smoothing of the input signal but also increases the computational memory required. For this work, we choose a Wiener filter of order $n=100$ as it provides a reasonable balance between smoothing and efficiency. The timeseries and frequency spectrum after applying the Wiener filter are shown in the sixth row in Fig.~\ref{fig:op_mic_results}.
We see a significant improvement in the timeseries of the recovered signal, however, a strong noise hum persists, audibly.

### 2021-11-01 paper/ifo-wanderingTone.tex
removing our comments to share source files
changing eqn to eq for AJP style guide

\begin{figure}
\begin{center}
\includegraphics[width=0.48\textwidth]{figures/viterbiDiagramSizes.pdf}
\caption{\label{fig:viterbi}
Schematic diagram of the Viterbi algorithm. 
The circular nodes represent elements in a time-frequency grid, labeled $t_0$ to $t_{N_t}$ (left to right) in time and $f_0$ to $f_{N_f}$ (bottom to top) in frequency. 
The size of the nodes represents the likelihood of a signal being present in each time-frequency element. 
The small to large sizing corresponds to low to high likelihood values in arbitrary units for this schematic. 
The objective is to find the most probable path of a signal through the grid from left to right.
All possible paths are shown by the lines. 
The black lines show the best path to each node and the gray lines show rejected paths. 
Some routes through the grid result in dead-ends with respect to optimally reaching the other side, such as the path ending at $t_1$ marked (a).
At $t=t_{N_t}$ the algorithm chooses the terminating frequency node which has the highest value given by Eqn.~\ref{eqn:pathprobability}, marked (b). 
The Viterbi path is the path leading to this node, highlighted by the thick-black lines, from (b) to the start at (c). 
}
\end{center}
\end{figure}

@@

Before we begin analyzing the data, we have no prior knowledge as to which frequency bin the signal starts in (i.e. the prior is flat between the minimum and maximum frequency bins).
At the first stage of the analysis, we define the probability of the system having frequency $f_j$ at the initial time $t_0$ to be equal to the (normalized) detection statistic of that state (i.e.\  ${\rm Pr}[f(t_0) = f_j] = F[t_0,f_j]$).
A specific path (which may not be optimal) is written as $f(t_0), f(t_1),\mathellipsis, f(t_n)$. 
The probability of a specific path given the data is 
\begin{eqnarray}
{\rm Pr}[ f(t_0), f(t_1),&&\mathellipsis, f(t_n) | {\rm data}] \nonumber\\
          && =F[t_n,f(t_n)] A[f(t_n),f(t_{n-1})] \nonumber \\
          && \times\mathellipsis\times A[f(t_1),f(t_0)] F[t_0,f(t_0)] \,.
\label{eqn:pathprobability}
\end{eqnarray}
The path $f^\ast(t_0),\mathellipsis,f^\ast(t_n)$ that maximises ${\rm Pr}[ f(t_0), f(t_1),\mathellipsis, f(t_n) | {\rm data}]$ is the optimal path terminating in the frequency bin $f^\ast(t_n)$. 
We note that the left-hand-side and right-hand-side of Eqn.~\ref{eqn:pathprobability} are both evaluated for a specific path and then we maximize over all such paths to find the optimal Viterbi path.

@@

\item Moving to time $t_2$, again we select the path which maximizes Eqn.~\ref{eqn:pathprobability} for each $f_j$. 
These paths are again shown by the solid black lines between the nodes at $t_1$ and $t_2$ in Fig.~\ref{fig:viterbi}.
Rejected paths are again shown by gray lines.

@@

\item We have now found the most probable path to each $f_j$ at $t=t_{N_t}$ and its probability (Eqn.~\ref{eqn:pathprobability}). 
We select the terminating frequency bin $f(t_{N_t})$ with the highest probability labeled as (b) in Fig.~\ref{fig:viterbi}.

@@

In continuous gravitational wave searches, the signal amplitude is expected to be small in comparison to the noise and its frequency can change unpredictably over time. % how to hyphenate ``continuous gravitational wave searches''?
The Viterbi algorithm's strength lies in its ability to track such signals through the data even in the presence of comparatively loud noise fluctuations in the time-frequency bins.
In the following section, we present the results of using the Viterbi algorithm with the table-top interferometer data.

>>>>

\begin{figure}
\begin{center}
\includegraphics[width=0.48\textwidth]{figures/viterbiDiagramSizes.pdf}
\caption{\label{fig:viterbi}
Schematic diagram of the Viterbi algorithm. 
The circular nodes represent elements in a time-frequency grid, labeled $t_0$ to $t_{N_t}$ (left to right) in time and $f_0$ to $f_{N_f}$ (bottom to top) in frequency. 
The size of the nodes represents the likelihood of a signal being present in each time-frequency element. 
The small to large sizing corresponds to low to high likelihood values in arbitrary units for this schematic. 
The objective is to find the most probable path of a signal through the grid from left to right.
All possible paths are shown by the lines. 
The black lines show the best path to each node and the gray lines show rejected paths. 
Some routes through the grid result in dead-ends with respect to optimally reaching the other side, such as the path ending at $t_1$ marked (a).
At $t=t_{N_t}$ the algorithm chooses the terminating frequency node which has the highest value given by Eq.~\ref{eqn:pathprobability}, marked (b). 
The Viterbi path is the path leading to this node, highlighted by the thick-black lines, from (b) to the start at (c). 
}
\end{center}
\end{figure}

@@

Before we begin analyzing the data, we have no prior knowledge as to which frequency bin the signal starts in (i.e. the prior is flat between the minimum and maximum frequency bins).
At the first stage of the analysis, we define the probability of the system having frequency $f_j$ at the initial time $t_0$ to be equal to the (normalized) detection statistic of that state (i.e.\  ${\rm Pr}[f(t_0) = f_j] = F[t_0,f_j]$).
A specific path (which may not be optimal) is written as $f(t_0), f(t_1),\mathellipsis, f(t_n)$. 
The probability of a specific path given the data is 
\begin{eqnarray}
{\rm Pr}[ f(t_0), f(t_1),&&\mathellipsis, f(t_n) | {\rm data}] \nonumber\\
          && =F[t_n,f(t_n)] A[f(t_n),f(t_{n-1})] \nonumber \\
          && \times\mathellipsis\times A[f(t_1),f(t_0)] F[t_0,f(t_0)] \,.
\label{eqn:pathprobability}
\end{eqnarray}
The path $f^\ast(t_0),\mathellipsis,f^\ast(t_n)$ that maximises ${\rm Pr}[ f(t_0), f(t_1),\mathellipsis, f(t_n) | {\rm data}]$ is the optimal path terminating in the frequency bin $f^\ast(t_n)$. 
We note that the left-hand-side and right-hand-side of Eq.~\ref{eqn:pathprobability} are both evaluated for a specific path and then we maximize over all such paths to find the optimal Viterbi path.

@@

\item Moving to time $t_2$, again we select the path which maximizes Eq.~\ref{eqn:pathprobability} for each $f_j$. 
These paths are again shown by the solid black lines between the nodes at $t_1$ and $t_2$ in Fig.~\ref{fig:viterbi}.
Rejected paths are again shown by gray lines.

@@

\item We have now found the most probable path to each $f_j$ at $t=t_{N_t}$ and its probability (Eq.~\ref{eqn:pathprobability}). 
We select the terminating frequency bin $f(t_{N_t})$ with the highest probability labeled as (b) in Fig.~\ref{fig:viterbi}.

@@

In continuous gravitational wave searches, the signal amplitude is expected to be small in comparison to the noise and its frequency can change unpredictably over time. 
The Viterbi algorithm's strength lies in its ability to track such signals through the data even in the presence of comparatively loud noise fluctuations in the time-frequency bins.
In the following section, we present the results of using the Viterbi algorithm with the table-top interferometer data.
